Select-based random access to variable-byte encodings

### 2020-02-26 gradu/msc_english_contents.tex
mods

One of the leading methods of data compression is variable-length coding \citep{Sal99}, where frequent sequences of data
are represented with shorter codewords. Because the sequences of data have different lenghts when compressed, it is 
not trivial to determine the exact location of a certain element. If this is required, the usual data compression algorithms are
inefficient. Fortunately this is not a requirement compression algorithms usually need to fulfill.

However, random access of compressed data is needed in compressed data structures. In most compression methods, the only way 
to this is to decompress data from the beginning. An integer compressing method with fast random access is explained later and compared
existing state-of-the-art methods.

@@

Variable-byte (VB) encoding \citep{Wil99} is a method of compressing integers via omitting leading zero bits. In normal data sets, it loses in 
compression performance to generic algorithms like Huffman encoding or Lempel-Ziv encoding, but in comparison it's faster to decode \citep{Bri09} 
and sometimes preferred due to it's simplicity. allows constant time random access. This is later explained in (TODO: link to later chapter).

@@

VB encoding splits each integer into blocks of $b$ bits and adds a continuation bit to the block to form chunks of length $1+b$. The extra bit is set to 1 only
on the block containing the least significant bits of the integer. This information is used in decoding to signal if next chunk continues the current 
integer. For example, let's assume $b = 4$ and let $n = 42$ be a 16 bit unsigned integer. The standard 8-bit representation of $n$ is 
\texttt{00000000 00101000}. When split to blocks of $b$ bits, it becomes \texttt{0000 0000 0010 1000}. Empty blocks are omitted and continuation bits 
are added to the remaining blocks. The result is \texttt{00010 11000}, which is the compressed data.

\medskip
\noindent\begin{minipage}{.5\textwidth}
\begin{algorithmic}
\Function{VBEncodeNumber}{$n$}
\State $bytes\gets $ list
\While{true}
\State $bytes$.prepend($n \bmod 128$)
\If{$n < 128$} \State break \EndIf
\State $n\gets n$ div $128$
\EndWhile
\State $bytes$.last() += $128$
\State \textbf{return} bytes
\EndFunction
\captionof{figure}{VByte encoding} \label{vbyte_enc}
\end{algorithmic}
\end{minipage}%
\begin{minipage}{.5\textwidth}
\begin{algorithmic}
\Function{VBEncode}{$numbers$}
\State $bytestream\gets $ list
\ForEach {$n \in numbers$}
\State $bytes \gets$ VBEncodeNumber($n$)
\State $bytestream$.extend($bytes$)
\EndFor
\State \textbf{return} bytestream
\EndFunction
\end{algorithmic}
\end{minipage}
\medskip

@@

\chapter{Directly addressable codes?}

@@

 - do cachegrind
 - 
 - use array instead of vector<uint8 t> for bris?
- comparison to basic implementation + Bri09
  - compare with b=2,4,8

@@

>>>>

One of the leading methods of data compression is variable-length coding \citep{Sal99}, where frequent sequences of data
are represented with shorter codewords. Because the sequences of data have different lengths when compressed, it is 
not trivial to determine the exact location of a certain element in the compressed data. If this is required, the usual 
data compression algorithms are inefficient. Fortunately this is not a requirement compression algorithms usually need to fulfill. However, 
random access of compressed data is useful in compressed data structures. It saves storage space, bandwith and increases the likelihood
of data already being in cache \citep{Sch02}. In most compression methods, the only way to do this is to decompress data from the beginning.

A variable-byte encoding based integer compression method with fast random access was first introduced by \citep{Bri09}. They used clever block 
reorganizing and $rank$ data structure to achieve random access. Their solution is currently the only published solution for the problem and it 
has been widely adopted.

An alternative solution with $select$ data structure is shown later. (TODO: refer to results)

--contributions?

@@

Variable-byte (VB) encoding \citep{Wil99} is a method for compressing integers via omitting leading zero bits that would be present in a longer fixed 
width word. In normal data sets, VB encoding loses in compression performance to generic algorithms like Huffman encoding or Lempel-Ziv encoding, but 
is generally faster to decode \citep{Bri09} and sometimes preferred due to its simplicity. As later shown in Chapter~\ref{chapter:DAC}, constant time 
random access is possible with VB encoding.

@@

Variable-byte encoding originates from and is used in MIDI music file \citep{Mid96} and several applications have a similar implementation of VB. Apache 
Lucene has vInt datatype. Wireless Application Protocol has a variable length unsigned integer uintvar, Google Protocol Buffers has a Base 128 Varint \citep{GooPB},
 Microsoft .Net framework offers "7BitEncodedInt" in BinaryReader and BinaryWriter classes and IBM DB2 has a variable byte  \citep{Bha09}.

Elias Delta and Gamma codes \citep{Eli75} are popular encoding methods for forementioned data sets. They assign short bit array codes to small numbers, 
which allows them to outperform VB encoding on small number datasets \Citep{Wil99}. Elias Delta and Gamma codes can't support fast random access efficiently because
of their changing code length.

VB encoding splits each integer into blocks of $b$ bits and adds a continuation bit to the block to form chunks of length $1+b$. The extra bit is set to 1 only
on the block containing the least significant bits of the integer. This information is used in decoding to signal if next chunk continues the current 
integer. For example, let's assume $b = 4$ and let $n = 42$ be a 16 bit unsigned integer. The standard 8-bit representation of $n$ is 
\texttt{00000000 00101000}. When split to blocks of $b$ bits, it becomes \texttt{0000 0000 0010 1000}. Empty blocks are omitted and continuation bits 
are added to the remaining blocks. The result is \texttt{00010 11000}, which is the compressed data.
\begin{figure}[ht]
\begin{algorithmic}
\Function{VBEncodeNumber}{$n$}
\State $bytes\gets $ list
\While{true}
\State \Call{prepend}{bytes,$n \bmod 128$}
\If{$n < 128$} \State break \EndIf
\State $n\gets n$ div $128$
\EndWhile
\State $bytes$[\Call{len}{$bytes$}-1] += $128$
\State \textbf{return} bytes
\EndFunction

\Function{VBEncode}{$numbers$}
\State $bytestream\gets $ list
\ForEach {$n \in numbers$}
\State $bytes \gets$ VBEncodeNumber($n$)
\State $bytestream$.extend($bytes$)
\EndFor
\State \textbf{return} bytestream
\EndFunction

\end{algorithmic}
\captionof{figure}{VByte encoding} \label{vbyte_enc}
\end{figure}

@@

\chapter{Directly addressable codes} \label{chapter:DAC}

@@

 - Run with 32 bit values, see what happens (some operations may be 32b?)
 - do cachegrind
 - 
 - use array instead of vector<uint8 t> for bris?
- comparison to basic implementation + Bri09
  - compare with b=2,4,8

@@

 - $rank$ vs rank? should formulas have a specific style?
